---
title: "class06"
author: "Selma Cifric (PID A69042976)"
format: pdf
toc: true
---
All functions in R have at least 3 things:
- A **name**, **arguments**, and **body**

#First silly function

```{r}
add <- function(x, y=1) {
  x + y
}
```

now, call function

```{r}
add(c(10, 10), 100)
```

## Second function

Write a function to generat erandom nucleotide sequences of a user specific length:

The `sample()` function can be helpful here. Replace within the function = TRUE vs FALSE

```{r}
v <- sample(c("A", "C", "G", "T"), size=50, replace = TRUE)
```

I want a 1 element long character vector that looks like a string and not "a" "c" "t".. but ACT. We use `paste` function where collapse="" will pull the elements into a string. I could also do collapse="--SC--" and it would use this as my spacings


```{r}

paste(v, collapse = "")
```

Generate a function to combine both of codes above

```{r}
generate_dna <- function(size=5) {
  v <- sample(c("A", "C", "G", "T"), size=size, replace = TRUE)
  paste(v, collapse = "")
}
```


```{r}
generate_dna(60)
```

IF and ELSE functions

```{r}
fasta <- TRUE
if(fasta) {
  cat("HELLO You!")
} else {
  cat("No you dont!")
}

```

Add the ability to return a multi-element vector or a single element fasta like vector.

```{r}
generate_fasta <- function(size=50, fasta=TRUE) {
  v <- sample(c("A", "C", "G", "T"), size=size, replace = TRUE)
  s <- (paste(v, collapse = ""))
  
  if(fasta) {
    return(s)
  } else {
    return(v)
  }
}

```

```{r}
generate_fasta(100, FALSE)
```

## A protein generating function

```{r}

generate_protein <- function(size=50, fasta=TRUE) {
  v <- sample(c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", 
                "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V"), 
              size=size, replace=TRUE)
  s <- paste(v, collapse = "")
  
  if(fasta) {
    return(s)
  } else {
    return(v)
  }
}

```

```{r}
generate_protein(6)
```

Use our new `generate_protein()` function to make random protein sequences of lenght 6-12 (i.e. one length 6, one length 7, etc up to length 12)

One way is to:

```{r}

generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)

```

OR use FOR loop `for()` loop:

```{r}
lengths <-  6:12
lengths

```

```{r}
for(i in lengths) {
  cat(">", i, " ")
  aa <- generate_protein(i)
  cat(aa)
  cat("\n")
}
generate_protein()
```

A third, and better, way to solve this is to use the `apply()` family of functions, specifically the `sapply()` function in this case.

```{r}
help(sapply)

sapply(6:12, generate_protein)

```



